---
title: "Connectivity deep biosphere microbes"
author: "George Westmeijer"
date: "26/07/2021"
output:
  html_document:
    df_print: paged
---

```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
```

## Read files
The sequence table ("seqtab.tsv") is stored in long format. Samples with less than 1000 reads have been removed. The sample metadata ('smd.tsv') contains the environment of the sample (environment) and allows grouping by replicates. 

```{r read data, message=FALSE, warning=FALSE}
smd <- read_tsv('smd.tsv', col_types = cols(.default = col_character())) %>%
  # Some samples have been filtered out above
  mutate(env = factor(env, 
                      levels = c('Baltic surface seawater',
                                 'Baltic benthic seawater',
                                 'Upper sediment',
                                 'Lower sediment',
                                 'Upper soil groundwater',
                                 'Lower soil groundwater',
                                 'Meteoric groundwater',
                                 'Modern marine groundwater',
                                 'Old saline groundwater'))) %>%
  group_by(env) %>% add_tally() %>% ungroup()

tax <- read_tsv('taxonomy.tsv', col_types = cols(.default = col_character()))

seqtab <- read_tsv('seqtab.tsv', col_types = cols(.default = col_character(), count = col_integer())) %>%
  # Filter out samples with less than 1000 reads
  group_by(sample) %>% filter(sum(count) > 1000) %>%
  # Calculate relative abundance within sample
  mutate(relab = count / sum(count)) %>% ungroup() %>%
  # Average the relab within an environment by dividing by the total
  inner_join(smd, by = 'sample') %>%
  group_by(env) %>% mutate(relab = relab / sum(relab)) %>% ungroup() %>%
  select(seqid, sample, count, relab) -> seqtab

chemistry <- read_tsv('chemistry.txt', col_types = cols(.default = col_double(), replicate = col_character())) %>%
  mutate(Cl = Cl / 1000, EC = EC / 100) %>%
  gather(parameter, value, -1) %>% 
  na.omit()
```


Define themes to use throughout the manuscript


```{r set theme}
theme_tidy <- function() {
  theme(
    axis.title.x = element_text(colour = "black", size = 8),
    axis.title.y = element_text(colour = "black", size = 8),
    axis.text.x = element_text(colour = "black", size = 8),
    axis.text.y = element_text(colour = "black", size = 8),
    legend.text = element_text(colour = "black", size = 8),
    strip.background = element_blank(),
    strip.placement = "outside",
    strip.text = element_text(colour = "black", size = 8, hjust = 0.5),
    panel.grid = element_blank(),
    legend.position = "bottom",
    aspect.ratio = 1.0,
    plot.background = element_blank(),
    panel.background = element_blank(),
    panel.border = element_rect(colour = "black", size = 0.5, fill = NA),
    legend.title = element_blank(),
    legend.box.background = element_blank(),
    legend.key = element_blank()
  )
}

theme_barplot <- function() {
  theme(
    axis.title.x = element_text(colour = "black", size = 8),
    axis.title.y = element_text(colour = "black", size = 8),
    axis.text.x = element_text(colour = "black", size = 8),
    axis.text.y = element_text(colour = "black", size = 8, margin = margin(0,-5,0,0, unit = 'mm')),
    legend.text = element_text(colour = "black", size = 8),
    strip.background = element_blank(),
    panel.grid = element_blank(),
    legend.position = "bottom",
    aspect.ratio = 0.7,
    strip.text = element_text(colour = "black", size = 8, hjust = 0.5),
    plot.background = element_blank(),
    panel.background = element_blank(),
    legend.title = element_blank(),
    legend.box.background = element_blank(),
    legend.key = element_blank(),
    legend.spacing.x = unit(0.5, 'mm'),
    panel.border = element_blank(),
    legend.key.size = unit(4.5, "mm"),
    legend.box.spacing = unit(1, 'mm'),
    axis.ticks.y = element_blank()
  )
}
```


```{r barplot function}
barplot <- function(asvs = seqtab, meta = smd, taxonomy = tax, taxlevel = "phylum", n = 11) {
  if (taxlevel == "phylum") {  
  asvs %>%
  inner_join(meta, by = 'sample') %>%
  inner_join(taxonomy, by = "seqid") %>%
  group_by(phylum, env) %>%
  # Sum the abundance of each phylum within a sample
  summarise(relab = sum(relab), .groups = 'drop_last') %>%
  # Calculate the mean abundance of each phylum over the categories
  summarise(mean_relab = sum(relab), .groups = 'drop') %>%
  filter(!is.na(phylum) & phylum != "uncultured bacterium") %>%
  top_n(11, mean_relab) -> t

taxonomy %>%
  left_join(t %>% transmute(phylum, topphylum = phylum), by = "phylum") %>%
  replace_na(list("topphylum" = "Other")) -> taxref

asvs %>%
  inner_join(taxref, by = "seqid") %>% 
  inner_join(meta, by = 'sample') %>%
  # Summarize in order to have the sum for each category and topphylum
  group_by(topphylum, env) %>% 
  summarise(relab = sum(relab), .groups = 'drop') -> asvs.phylum
  return(asvs.phylum)
  }
  else if (taxlevel == "class") {
  asvs %>%
  inner_join(meta, by = 'sample') %>%
  inner_join(taxonomy, by = "seqid") %>%
  group_by(class, env) %>%
  # Sum the abundance of each phylum within a sample
  summarise(relab = sum(relab), .groups = 'drop_last') %>%
  # Calculate the mean abundance of each phylum over the categories
  summarise(mean_relab = sum(relab), .groups = 'drop') %>%
  filter(!is.na(class) & class != "uncultured bacterium") %>%
  top_n(11, mean_relab) -> t

taxonomy %>%
  left_join(t %>% transmute(class, topphylum = class), by = "class") %>%
  replace_na(list("topphylum" = "Other")) -> taxref

asvs %>%
  inner_join(taxref, by = "seqid") %>% 
  inner_join(meta, by = 'sample') %>%
  # Summarize in order to have the sum for each category and topphylum
  group_by(topphylum, env) %>% 
  summarise(relab = sum(relab), .groups = 'drop') -> asvs.class
  return(asvs.class)
  }
  else {
  asvs %>%
  inner_join(meta, by = 'sample') %>%
  inner_join(taxonomy, by = "seqid") %>%
  group_by(order, env) %>%
  # Sum the abundance of each phylum within a sample
  summarise(relab = sum(relab), .groups = 'drop_last') %>%
  # Calculate the mean abundance of each phylum over the categories
  summarise(mean_relab = sum(relab), .groups = 'drop') %>%
  filter(!is.na(order) & !grepl(pattern = "uncultured", x = order)) %>%
  top_n(n, mean_relab) -> t

taxonomy %>%
  left_join(t %>% transmute(order, topphylum = order), by = "order") %>%
  replace_na(list("topphylum" = "Other")) -> taxref

asvs %>%
  inner_join(taxref, by = "seqid") %>% 
  inner_join(meta, by = 'sample') %>%
  # Summarize in order to have the sum for each category and topphylum
  group_by(topphylum, env) %>% 
  summarise(relab = sum(relab), .groups = 'drop') -> asvs.order
  return(asvs.order)
  }
}
```

## Alpha diversity using Shannon-Weaver index


Calculate Shannon-Weaver for each sample and combine with sample metadata


```{r estimate alpha diversity}
seqtab %>%
  select(-relab) %>%
  spread(seqid, count, fill = 0) %>% column_to_rownames('sample') %>%
  vegan::diversity() %>% as.data.frame() %>% gather(seqid, shannon, -1) %>%
  rename(shannon = 1) %>%
  rownames_to_column('sample') %>%
  inner_join(smd, by = "sample") %>%
  group_by(replicate) %>%
  mutate(meandiv = mean(shannon)) %>% ungroup() %>%
  arrange(env) -> adiv
```


Plot alpha diversity


```{r plot alpha diversity, echo = FALSE}
adiv %>%
  ggplot(aes(x = shannon, y = fct_rev(env))) + 
  geom_boxplot() +
  geom_jitter(width = 0.01, alpha = 0.4, height = 0.1) +
  labs(x = "Alpha diversity (Shannon index)", y = "") +
  theme_tidy() + theme(aspect.ratio = 1.8)
```


```{r export alpha diversity plot}
ggsave("figures/fig_3a.pdf", width = 10, height = 10, units = "cm")
```


```{r chemistry and alpha}
chemistry %>%
  filter(parameter %in% c('O18', 'ph', 'depth', 'Cl', 'DOC', 'FEII')) %>%
  inner_join(smd, by = 'replicate') %>%
  mutate(parameter = factor(parameter, levels = c("depth","DOC","O18","Cl","FEII","ph"))) %>%
  group_by(env, parameter) %>% 
  mutate(env = sub("groundwater", "\ngroundwater", env)) %>%
  ggplot(aes(x = fct_rev(fct_relevel(env, c("Upper soil \ngroundwater"))), y = value)) +
  geom_boxplot(na.rm = TRUE) +
  geom_jitter(width = 0.01, alpha = 0.4, height = 0.1, na.rm = TRUE) +
  coord_flip() +
  facet_wrap(~parameter, scales = 'free_x', nrow = 1, strip.position = 'bottom', labeller = 
               as_labeller(c(
                 depth = "Depth~(m)", 
                 DOC = "DOC~(mg~L^-1)", 
                 Cl = 'Cl^{-phantom()}~(g~L^-1)',
                 FEII = 'Fe^{2+phantom()} ~ (mg~L^-1)',
                 O18 = 'delta^{18}*O~(ppt)', 
                 ph = 'pH'), default = label_parsed)) +
  theme_tidy() + theme(aspect.ratio = 3.0, 
                       strip.placement = "outside", 
                       axis.title.x = element_blank(), 
                       axis.title.y = element_blank())
```


```{r export chemistry plot}
ggsave("figures/fig_1.pdf", width = 20, height = 10, units = "cm")
```


```{r supplemental chemistry}
chemistry %>%
  filter(parameter %in% c('depth','Mn', 'Mg', 'SO4', 'Sr', 'EC')) %>%
  inner_join(smd, by = 'replicate') %>%
  mutate(parameter = factor(parameter, levels = c("depth","EC","Sr","Mn","Mg","SO4"))) %>%
  group_by(env, parameter) %>%
  mutate(env = sub("groundwater", "\ngroundwater", env)) %>%
  ggplot(aes(x = fct_rev(fct_relevel(env, c("Upper soil \ngroundwater"))), y = value)) +
  geom_boxplot(na.rm = TRUE) +
  geom_jitter(width = 0.01, alpha = 0.4, height = 0.1, na.rm = TRUE) +
  coord_flip() +
  facet_wrap(~parameter, scales = 'free_x', nrow = 1, strip.position = 'bottom', labeller = 
               as_labeller(c(
                 depth = "Depth~(m)", 
                 EC = "EC~(mS~cm^-1)", 
                 Sr = 'Sr~(mg~L^-1)',
                 Mn = 'Mn~(mg~L^-1)',
                 Mg = 'Mg~(mg~L^-1)', 
                 SO4 = 'SO[4]^{2-phantom()}~(mg~L^-1)'), default = label_parsed)) +
 theme_tidy() + theme(aspect.ratio = 3.0, 
                       strip.placement = "outside", 
                       axis.title.x = element_blank(), 
                       axis.title.y = element_blank())
```


```{r export supplemental chemistry}
ggsave("figures/fig_s1.pdf", width = 20, height = 10, units = "cm")
```


```{r alpha diversity correlation}
chemistry %>%
  group_by(replicate, parameter) %>% summarise(value = mean(value), .groups = "drop") %>%
  filter(parameter %in% c("depth","DOC")) %>%
  inner_join(adiv, by = "replicate") %>%
  ggplot(aes(x = value, y = meandiv)) +
  geom_point(aes(color = env), size = 3, alpha = 0.7, na.rm = TRUE) +
  geom_smooth(method = "lm", formula = y ~ x,color = "black", fill = "#d9d9d9", se = TRUE, size = 0.8) +
  labs(x = "", y = 'Alpha diversity (Shannon index)') +
  scale_color_manual(values = c("#fb9a99","#e31a1c","#fdbf6f","#b15928","#cab2d6")) +
  facet_wrap(~parameter, scales = 'free_x', nrow = 1, strip.position = 'bottom', labeller = 
               as_labeller(c(
                 depth = "Depth~(m)", 
                 DOC = "Dissolved~organic~carbon~(mg~L^-1)"), default = label_parsed)) +
  theme_tidy() + guides(color = guide_legend(nrow = 2)) + theme(legend.box.spacing = unit(-3, 'mm'))
```


```{r export correlation plot}
ggsave("figures/fig_3bc.pdf", width = 14, height = 10, units = "cm")
```


```{r pearson correlation on alpha diversity}
# Correlation alpha diversity and depth
chemistry %>%
  group_by(replicate, parameter) %>% summarise(value = mean(value), .groups = "drop") %>%
  filter(parameter == "depth") %>%
  inner_join(adiv, by = "replicate") %>%
  stats::cor.test(x = .$meandiv, y = .$value, method = "pearson", alternative = "two.sided")

# Correlation alpha diversity and dissolved organic carbon
chemistry %>%
  group_by(replicate, parameter) %>% summarise(value = mean(value), .groups = "drop") %>%
  filter(parameter == "DOC") %>%
  inner_join(adiv, by = "replicate") %>%
  stats::cor.test(x = .$meandiv, y = .$value, method = "pearson", alternative = "two.sided")

# Correlation alpha diversity and chloride concentration
chemistry %>%
  group_by(replicate, parameter) %>% summarise(value = mean(value), .groups = "drop") %>%
  filter(parameter == "Cl") %>%
  inner_join(adiv, by = "replicate") %>%
  stats::cor.test(x = .$meandiv, y = .$value, method = "pearson", alternative = "two.sided")
```
  

Statistical testing on alpha diversity, start with testing for the assumptions. Use the mean over the replicates (meandiv) for statistical testing.

```{r anova on alpha diversity}
suppressPackageStartupMessages(library(car))
# Normality of the data (null hypothesis data is normally distributed)
shapiro.test(adiv$shannon)
# Homogeneity of variance using Levene (null hypothesis variance is equal)
leveneTest(shannon ~ env, data = adiv)
# Non-parametric testing (null hypothesis diversity is equal)
kruskal.test(meandiv ~ env, data = adiv)
# Run ANOVA (null hypothesis diversity is equal)
aov(meandiv ~ env, data = adiv) %>% summary()
# Post-hoc using Tukey (null hypothesis diversity is equal)
aov(meandiv ~ env, data = adiv) %>% TukeyHSD(which = "env")
```

```{r quantile-quantile plot}
qqPlot(adiv$shannon, ylab = "Alpha diversity (Shannon index)")
```


## Plot sequence length distribution among the different environments

```{r sequence length distribution, echo = FALSE}
seqs <- read_tsv('sequences.tsv', col_types = cols(.default = col_character()))

seqtab %>%
  inner_join(smd, by = "sample") %>%
  inner_join(seqs, by = 'seqid') %>%
  ggplot(aes(x = env, y = nchar(sequence))) +
  geom_boxplot() +
  labs(x = '', y = 'ASV length (nt)') +
  theme_bw() +
  theme(axis.text.x.bottom = element_text(angle = 90, hjust = 1.0))
```

## Ordination using NMDS, filtering out ASVs with a count less than 2

```{r run nmds, results='hide'}
seqtab %>%
  group_by(seqid) %>% filter(count > 2) %>% ungroup() %>%
  select(seqid, sample, count) %>% spread(seqid, count, fill = 0) %>%
  column_to_rownames("sample") -> t

set.seed(999)
t %>%  vegan::metaMDS(trymax = 20, k = 2, autotransform = T) -> nmds
```

Extract the scores

```{r nmds.scores}
vegan::scores(nmds) %>%
  as.data.frame() %>%
  rownames_to_column("sample") %>%
  inner_join(smd, by = "sample") -> nmds.scores
```

Plot the ordination

```{r nmds plot, include = T}
nmds.scores %>%
  ggplot(aes(x = NMDS1, y = NMDS2)) +
  geom_point(aes(colour = env), size = 4.0, alpha = 0.7) +
  scale_color_manual(values = 
                       c("#a6cee3","#1f78b4","#b2df8a","#33a02c",
                         "#fb9a99","#e31a1c","#fdbf6f","#b15928","#cab2d6")) +  
  labs(x = "", y = "") +
  geom_vline(xintercept = 0, linetype = 'dotted') +
  geom_hline(yintercept = 0, linetype = 'dotted') +
  annotate('text', x = -2.3, y = -2.6, size = 3, label = paste('stress = ', round(nmds$stress, digits = 2))) +
  theme_tidy() + guides(color = guide_legend(nrow = 3)) + 
  theme(aspect.ratio = 0.8, legend.spacing.x = unit(-0.5, 'mm'))
```

```{r export nmds}
ggsave("figures/fig_4.pdf", width = 12, height = 12, units = "cm")
```

```{r nmds deep biosphere groundwaters}
seqtab %>%
  filter(count > 4 & sample %in% smd[smd$surface == 'no', ]$sample) %>%
  select(seqid, sample, count) %>% spread(seqid, count, fill = 0) %>%
  column_to_rownames("sample") -> t

set.seed(99)
t %>%  vegan::metaMDS(trymax = 50, k = 2, autotransform = T) -> nmds.groundwater
```

```{r plot nmds deep biosphere groundwaters}
vegan::scores(nmds.groundwater) %>%
  as.data.frame() %>%
  rownames_to_column("sample") %>%
  inner_join(smd, by = "sample") %>%
  ggplot(aes(x = NMDS1, y = NMDS2)) +
  geom_point(aes(colour = env), size = 4.0, alpha = 0.7) +
  scale_color_brewer(palette = 'Paired') +
  labs(x = "", y = "") +
  geom_vline(xintercept = 0, linetype = 'dotted') +
  geom_hline(yintercept = 0, linetype = 'dotted') +
  annotate('text', x = -4, y = -2.4, size = 3, label = paste('stress = ', round(nmds.groundwater$stress, digits = 2))) +
  theme_tidy() + guides(color = guide_legend(nrow = 3)) + 
  theme(aspect.ratio = 0.8, legend.spacing.x = unit(-0.5, 'mm'), legend.position = c(0.20,0.13))
```

```{r export nmds deep biosphere groundwaters}
ggsave("figures/fig_s4.pdf", width = 12, height = 12, units = "cm")
```

Statistical testing on beta diversity

```{r statistical testing beta diversity}
seqtab %>%
  select(-relab) %>%
  inner_join(smd[, c("sample", "env", "replicate")], by = "sample") %>%
  spread(seqid, count, fill = 0) %>%
  column_to_rownames('sample') -> df

pairwiseAdonis::pairwise.adonis2(df[,3:ncol(df)] ~ env, data = df)
```


Test the homogeneity of group variation

```{r test homogeneity of within-group variation}
seqtab %>%
  select(-relab) %>% spread(seqid, count, fill = 0) %>%
  column_to_rownames("sample") %>%
  vegan::vegdist() %>%
  vegan::betadisper(sqrt.dist = T, group = smd %>% arrange(sample) %>% pull(env)) %>%
  anova()
```

## Bar plots

```{r plot bar plot phylum}
x.label = smd %>% select(env, n) %>% distinct() %>% mutate(env = paste(env, n, sep = "\nn = ")) %>% pull(env)

barplot() %>%
  # Call the plot
  ggplot(aes(x = fct_rev(env), y = relab, fill = fct_relevel(topphylum, c("Other")))) +
  labs(x = '', y = 'Relative abundance') +
  geom_col() +
  scale_fill_brewer(palette = 'Paired') +
  coord_flip() + 
  scale_x_discrete(labels = rev(x.label)) +
  scale_y_continuous(trans = 'reverse', labels = c('1.0','0.75','0.50','0.25','0')) +
  theme_barplot()
```

```{r export bar plot}
ggsave("figures/fig_2a.pdf", height = 16, width = 16, units = "cm")
```

```{r barplot class taxonomic level}
barplot(taxlevel = "class") %>%
  # Call the plot
  ggplot(aes(x = fct_rev(env), y = relab, fill = fct_relevel(topphylum, c("Other")))) +
  labs(x = '', y = 'Relative abundance') +
  geom_col() +
  scale_fill_brewer(palette = 'Paired') +
  coord_flip() + 
  scale_y_continuous(trans = 'reverse', labels = c('1.0','0.75','0.50','0.25','0')) +
  theme_barplot() -> fig_s3a
```

```{r barplot order taxonomic level}
my.pal <- colorRampPalette(RColorBrewer::brewer.pal(12, "Paired"))
paired.pal <- my.pal(20)

barplot(taxlevel = "order", n = 19) %>%  
  mutate(topphylum = gsub("Candidatus", "Ca.", topphylum)) %>%
  # Call the plot
  ggplot(aes(x = fct_rev(env), y = relab, fill = fct_relevel(topphylum, c("Other")))) +
  labs(x = '', y = 'Relative abundance') +
  geom_col() + 
  scale_fill_manual(values = paired.pal) +
  coord_flip() + scale_y_continuous(trans = 'reverse', labels = c('1.0','0.75','0.50','0.25','0')) +
  theme_barplot() -> fig_s3b
```

```{r export barplot class plus order}
cowplot::plot_grid(fig_s3a, fig_s3b, labels = c('a)','b)'), nrow = 2, label_size = 12, label_fontface = "plain")
ggsave("figures/fig_s3.pdf", height = 20, width = 20, units = "cm")
```

## Core / unique community

The core community is defined as the ASVs that are present in each of the three groundwater types. The unique commnunity (accesory community in manuscript) are the taxa that occur in a groundwater type but are not present in the other two. By filtering "n == 3", sequences are selected that are present in all three groundwaters; by filtering "n == 1", only the ASVs are selected that are not present in multiple groundwaters.

```{r define core asvs}
seqtab %>%
  # Select subsurface ASVs
  inner_join(smd, by = 'sample') %>% filter(surface == 'no') %>%
  select(seqid, env) %>% 
  # Keep only unique ASVs in each environment
  distinct() %>%
  group_by(seqid) %>% add_tally() %>% ungroup() %>%
  # Require to be present in all groundwater types by stating n == 3
  mutate(community = if_else(n == 1, 'acc', 'core')) %>%
  # n == 2 is not of interest here
  filter(n != 2) %>% select(seqid, community) %>% distinct() -> core
```

```{r top phyla in core, message=FALSE, warning=FALSE}
seqtab %>%
  filter(seqid %in% core[core$community == 'core',]$seqid) %>%
  barplot() %>% filter(topphylum != "Other") %>%
  left_join(tax, by = c("topphylum" = "phylum")) %>% pull(seqid) -> t

seqtab %>%
  filter(seqid %in% core[core$community == 'core',]$seqid) %>%
  inner_join(tax, by = "seqid") %>%
  mutate(topphylum = if_else(seqid %in% t, phylum, "Other")) %>%
  inner_join(smd, by = 'sample') %>%
  filter(surface == 'no') %>%
  group_by(topphylum, env) %>% 
  summarise(relab = sum(relab), .groups = 'drop') %>%
  ggplot(aes(x = fct_rev(env), y = relab, fill = fct_relevel(topphylum, c("Other")))) +
  labs(x = '', y = '') +
  geom_col() + scale_fill_brewer(palette = 'Paired') +
  coord_flip() + theme_barplot() + theme(aspect.ratio = 1/4) -> core.phylum

seqtab %>%
  filter(seqid %in% core[core$community == 'core',]$seqid) %>%
  barplot(taxlevel = "class") %>% filter(topphylum != "Other") %>%
  left_join(tax, by = c("topphylum" = "class")) %>% pull(seqid) -> t

seqtab %>%
  filter(seqid %in% core[core$community == 'core',]$seqid) %>%
  inner_join(tax, by = "seqid") %>%
  mutate(topphylum = if_else(seqid %in% t, class, "Other")) %>%
  inner_join(smd, by = 'sample') %>%
  filter(surface == 'no') %>%
  group_by(topphylum, env) %>% 
  summarise(relab = sum(relab), .groups = 'drop') %>%
  ggplot(aes(x = fct_rev(env), y = relab, fill = fct_relevel(topphylum, c("Other")))) +
  labs(x = '', y = '') +
  geom_col() + scale_fill_brewer(palette = 'Paired') +
  coord_flip() + theme_barplot() + theme(aspect.ratio = 1/4) -> core.class

seqtab %>%
  filter(seqid %in% core[core$community == 'core',]$seqid) %>%
  barplot(taxlevel = "order") %>% filter(topphylum != "Other") %>%
  left_join(tax, by = c("topphylum" = "order")) %>% pull(seqid) -> t

seqtab %>%
  filter(seqid %in% core[core$community == 'core',]$seqid) %>%
  inner_join(tax, by = "seqid") %>%
  mutate(topphylum = if_else(seqid %in% t, order, "Other")) %>%
  inner_join(smd, by = 'sample') %>%
  filter(surface == 'no') %>%
  group_by(topphylum, env) %>% 
  summarise(relab = sum(relab), .groups = 'drop') %>%
  mutate(topphylum = gsub("Candidatus", "Ca.", topphylum)) %>%
  ggplot(aes(x = fct_rev(env), y = relab, fill = fct_relevel(topphylum, c("Other")))) +
  labs(x = '', y = '') +
  geom_col() + scale_fill_brewer(palette = 'Paired') +
  coord_flip() + theme_barplot() + theme(aspect.ratio = 1/4) -> core.order

cowplot::plot_grid(core.phylum, core.class, core.order, nrow = 3, labels = c("a)", "", ""), label_fontface = "plain", label_size = 12)
ggsave("figures/fig_s5b.pdf", width = 16, height = 18, units = "cm")
```

```{r top phyla in acc, message=FALSE, warning=FALSE}
seqtab %>%
  filter(seqid %in% core[core$community == 'acc',]$seqid) %>%
  barplot() %>% filter(topphylum != "Other") %>%
  left_join(tax, by = c("topphylum" = "phylum")) %>% pull(seqid) -> t

seqtab %>%
  filter(seqid %in% core[core$community == 'acc',]$seqid) %>%
  inner_join(tax, by = "seqid") %>%
  mutate(topphylum = if_else(seqid %in% t, phylum, "Other")) %>%
  inner_join(smd, by = 'sample') %>%
  filter(surface == 'no') %>%
  group_by(topphylum, env) %>% 
  summarise(relab = sum(relab), .groups = 'drop') %>%
  ggplot(aes(x = fct_rev(env), y = relab, fill = fct_relevel(topphylum, c("Other")))) +
  labs(x = '', y = '') +
  geom_col() + scale_fill_brewer(palette = 'Paired') +
  coord_flip() + theme_barplot() + theme(aspect.ratio = 1/4) -> acc.phylum

seqtab %>%
  filter(seqid %in% core[core$community == 'acc',]$seqid) %>%
  barplot(taxlevel = "class") %>% filter(topphylum != "Other") %>%
  left_join(tax, by = c("topphylum" = "class")) %>% pull(seqid) -> t

seqtab %>%
  filter(seqid %in% core[core$community == 'acc',]$seqid) %>%
  inner_join(tax, by = "seqid") %>%
  mutate(topphylum = if_else(seqid %in% t, class, "Other")) %>%
  inner_join(smd, by = 'sample') %>%
  filter(surface == 'no') %>%
  group_by(topphylum, env) %>% 
  summarise(relab = sum(relab), .groups = 'drop') %>%
  ggplot(aes(x = fct_rev(env), y = relab, fill = fct_relevel(topphylum, c("Other")))) +
  labs(x = '', y = '') +
  geom_col() + scale_fill_brewer(palette = 'Paired') +
  coord_flip() + theme_barplot() + theme(aspect.ratio = 1/4) -> acc.class

seqtab %>%
  filter(seqid %in% core[core$community == 'acc',]$seqid) %>%
  barplot(taxlevel = "order") %>% filter(topphylum != "Other") %>%
  left_join(tax, by = c("topphylum" = "order")) %>% pull(seqid) -> t

seqtab %>%
  filter(seqid %in% core[core$community == 'acc',]$seqid) %>%
  inner_join(tax, by = "seqid") %>%
  mutate(topphylum = if_else(seqid %in% t, order, "Other")) %>%
  inner_join(smd, by = 'sample') %>%
  filter(surface == 'no') %>%
  group_by(topphylum, env) %>% 
  summarise(relab = sum(relab), .groups = 'drop') %>%
  mutate(topphylum = gsub("Candidatus", "Ca.", topphylum)) %>%
  ggplot(aes(x = fct_rev(env), y = relab, fill = fct_relevel(topphylum, c("Other")))) +
  labs(x = '', y = '') +
  geom_col() + scale_fill_brewer(palette = 'Paired') +
  coord_flip() + theme_barplot() + theme(aspect.ratio = 1/4) -> acc.order

cowplot::plot_grid(acc.phylum, acc.class, acc.order, nrow = 3, labels = c("b)", "", ""), label_fontface = "plain", label_size = 12)
ggsave("figures/fig_s5b.pdf", width = 16, height = 18, units = "cm")
```

## Overlapping taxa

Which ASVs occur in multiple environments, also including surface environments. A small matrix is prepared containing all possible pairwise combinations between the different environments and this is piped to intersect().

```{r relative abundance within category}
seqtab %>%
  # Calculate the relab within environment
  inner_join(smd, by = 'sample') %>% 
  group_by(env, seqid) %>% 
  summarise(relab = sum(relab), .groups = 'drop') -> reference
```


```{r count overlapping taxa}
expand_grid(x = unique(smd$env), y = unique(smd$env)) %>%
  mutate(overlap = 
           apply(., 1, function(x) intersect(
    reference[reference$env == x[1], "seqid"], 
    reference[reference$env == x[2], "seqid"]) %>% nrow())
    ) %>%
  mutate(abundance = 
           apply(., 1, function(x) intersect(
    reference[reference$env == x[1], "seqid"], 
    reference[reference$env == x[2], "seqid"]) %>% 
      inner_join(reference, by = "seqid") %>%
      filter(env == x[1]) %>%
      pull(relab) %>% 
      sum() %>% 
      round(digits = 2))
  )
```


```{r identivy asvs that are present in surface}
seqtab %>% 
  inner_join(smd, by = 'sample') %>%
  select(seqid, surface) %>% distinct() %>%
  group_by(seqid) %>% add_tally() %>% ungroup() %>%
  # By selecting n == 2, asvs are selected that are present in both surface and subsurface
  filter(n == 2) %>% pull(seqid) %>% unique() -> overlap
```


```{r plot overlap and non-overlapping community}
seqtab %>%
  filter(seqid %in% overlap) %>%
  barplot() %>% filter(topphylum != "Other") %>%
  left_join(tax, by = c("topphylum" = "phylum")) %>% pull(seqid) -> t

seqtab %>%
  filter(seqid %in% overlap) %>%
  inner_join(tax, by = "seqid") %>%
  mutate(topphylum = if_else(seqid %in% t, phylum, "Other")) %>%
  inner_join(smd, by = 'sample') %>%
  filter(surface == 'no') %>%
  group_by(topphylum, env) %>% 
  summarise(relab = sum(relab), .groups = 'drop') %>%
  ggplot(aes(x = fct_rev(env), y = relab, fill = fct_relevel(topphylum, c("Other")))) +
  labs(x = '', y = '') +
  geom_col() + scale_fill_brewer(palette = 'Paired') +
  coord_flip() + theme_barplot() + theme(aspect.ratio = 1/4) -> a

seqtab %>%
  filter(!seqid %in% overlap) %>%
  inner_join(tax, by = "seqid") %>%
  mutate(topphylum = if_else(seqid %in% t, phylum, "Other")) %>%
  inner_join(smd, by = 'sample') %>%
  filter(surface == 'no') %>%
  group_by(topphylum, env) %>% 
  summarise(relab = sum(relab), .groups = 'drop') %>%
  ggplot(aes(x = fct_rev(env), y = relab, fill = fct_relevel(topphylum, c("Other")))) +
  labs(x = '', y = '') +
  geom_col() + scale_fill_brewer(palette = 'Paired') +
  coord_flip() + theme_barplot() + theme(aspect.ratio = 1/4) -> b

cowplot::plot_grid(a + theme(legend.position = 'none'), b + theme(legend.position = 'none'), cowplot::get_legend(b), nrow = 3, 
                   labels = c('a)','b)'),
                   label_fontface = "plain")
ggsave("figures/fig_5bc.pdf", height = 12, width = 16, units = "cm")
```


## Pheatmap

```{r pheatmap}
seqtab %>%
  filter(seqid %in% overlap & sample %in% smd[smd$surface == 'no',]$sample) %>%
  inner_join(smd, by = 'sample') %>%
  inner_join(tax, by = "seqid") %>%
  group_by(seqid, env) %>%
  # Sum the abundance of each phylum within a sample
  summarise(relab = sum(relab), .groups = 'drop_last') %>%
  # Calculate the mean abundance of each phylum over the categories
  summarise(mean_relab = sum(relab), .groups = 'drop') %>%
  top_n(30, mean_relab) %>%
  inner_join(tax, by = 'seqid')-> t

t %>%
  # Mutate taxonomies that only say 'uncultured'
  mutate(across(c(class, order, family, genus, species), ~gsub('uncultured', NA, .))) %>%
  # Create a label based on the highest taxonomic level
  mutate(label = coalesce(genus, family, order, class, phylum, domain)) %>%
  # Add a unique ASV labl
  mutate(label = paste("ASV", sprintf('%02d', row_number()), label)) %>%
  select(label, seqid) %>%
  # Join with seqtab to obtain abundance per sample
  inner_join(seqtab, by = 'seqid') %>%
  # Join with metadata to add categories
  inner_join(smd, by = 'sample') %>%
  select(label, env, relab) %>%
  # Sum abundance within each grouping
  group_by(env, label) %>% summarise(relab = sum(relab), .groups = 'drop') %>%
  # Transform abundance
  mutate(relab = sqrt(relab)) %>%
  spread(env, relab, fill = 0) %>% column_to_rownames('label') %>%
  # Convert to a matrix
  as.matrix() -> heatmap_matrix

library(RColorBrewer)
pheatmap::pheatmap(heatmap_matrix,
                   filename = 'figures/pheatmap.pdf',
                   color = colorRampPalette(brewer.pal(8, "Blues"))(7), 
                   breaks = c(0, 0.001, 0.01, 0.02, 0.05, 0.1, 0.2, 0.35),
                   legend_breaks = c(0, 0.1, 0.2, 0.30), legend_labels = c('0','0.1','0.2','0.3'),
                   border_color = NA, fontsize = 9,
                   cellwidth = 15, cellheight = 15, 
                   scale = 'none', cluster_rows = T, cluster_cols = F, treeheight_row = 25,
                   clustering_distance_rows = 'euclidean')
```